---
title: "Multimodal NGS analysis of gene expression and chromatin state in olfactory epithelium"
subtitle: "Day 1 — Setting up your environment and downloading the data"
author: "Lisa Korneeva"
date: today
format:
  html:
    grid:
      body-width: 1000px
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    theme:
      - flatly
      - custom.scss
    highlight-style: github
    code-fold: show
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

---

# Overview

Welcome to the course! Today we will:

- Find and explore the dataset from the published paper
- Set up the **computational environment** on your server
- Download the raw **RNA-seq data** using **nf-core/fetchngs**
- Rename the downloaded files to meaningful sample names

By the end of the day you should have (hopefully) all raw data downloaded and renamed.

---

# 1. The Dataset

## 1.1 Paper

We will be working with a published dataset from a study on zone-specific gene expression
in the mouse main olfactory epithelium (MOE). Go to the following paper and read it,
focusing on the **Materials and Methods** section, specifically the part describing
the **RNA-seq** and **ChIP-seq** experiments:

> **Paper:** [https://elifesciences.org/articles/87445#s4](https://elifesciences.org/articles/87445#s4)

**Exercise:** Answer the following questions based on the Methods section:

1. What cell types were isolated and how were they sorted?
2. What sequencing platform(s) were used?
3. What read length was used for sequencing?
4. What library preparation kit was used?
5. Where is the raw data deposited? What is the GEO accession number?

## 1.2 Data on NCBI GEO

Go to NCBI GEO and search for the dataset using the accession number you found:

> **NCBI GEO:** [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi)

Browse the samples and identify the RNA-seq and ChIP-seq samples. We will be working with the following 8 samples for RNA-seq and 7 samples for ChIP-seq:

```{r sample-table, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

samples <- data.frame(
  `Sample Name` = c(
    "RNAseq_Zone1_Ngn1GFPdim_r1",
    "RNAseq_Zone1_Ngn1GFPdim_r2",
    "RNAseq_Zone1_Ngn1GFPdim_r3",
    "RNAseq_Zone5_Ngn1GFPdim_r1",
    "RNAseq_Zone5_Ngn1GFPdim_r2",
    "RNAseq_Zone5_Ngn1GFPdim_r3",
    "RNAseq_Zone2_Ngn1GFPdim_r1",
    "RNAseq_Zone2_Ngn1GFPdim_r2",
    "nchip_Z1_OMPgfp_H3K9me3_r1",
    "nchip_Z1_OMPgfp_H3K9me3_r2",
    "nchip_Z2_OMPgfp_H3K9me3_r1",
    "nchip_Z2_OMPgfp_H3K9me3_r2",
    "nchip_Z5_OMPgfp_H3K9me3_r1",
    "nchip_Z5_OMPgfp_H3K9me3_r2",
    "nchip_MOE_H3K9me3_r1"
  ),
  `GSM Accession` = c(
    "GSM4815274", "GSM4815275", "GSM4815276",
    "GSM4815285", "GSM4815286", "GSM4815287",
    "GSM4815294", "GSM4815295", "GSM4815328",
    "GSM4815333", "GSM4815330", "GSM4815334",
    "GSM4815332", "GSM4815335", "GSM4815345"
  ),
  `SRR Accession` = c(
    "SRR12761392", "SRR12761393", "SRR12761394",
    "SRR12761403", "SRR12761404", "SRR12761405",
    "SRR12761412", "SRR12761413", "SRR12761446",
    "SRR12761451", "SRR12761448", "SRR12761452",
    "SRR12761450", "SRR12761453", "SRR12761463"
  ),
  `Renamed to` = c(
    "RNA_Zone1_mOSN_Rep1", "RNA_Zone1_mOSN_Rep2", "RNA_Zone1_mOSN_Rep3", 
    "RNA_Zone5_mOSN_Rep1", "RNA_Zone5_mOSN_Rep2", "RNA_Zone5_mOSN_Rep3", 
    "RNA_Zone2_mOSN_Rep1", "RNA_Zone2_mOSN_Rep2", "ChIP_Zone1_mOSN_Rep1", 
    "ChIP_Zone1_mOSN_Rep2", "ChIP_Zone2_mOSN_Rep1", "ChIP_Zone2_mOSN_Rep2", 
    "ChIP_Zone5_mOSN_Rep1", "ChIP_Zone5_mOSN_Rep2", "ChIP_Bulk_mOSN_Rep1" 
  ),
  Technique = c(rep("RNA-seq", 8), rep("ChIP-seq", 7)), 
  Zone = c("Zone 1", "Zone 1", "Zone 1", 
           "Zone 5", "Zone 5", "Zone 5", 
           "Zone 2", "Zone 2", "Zone 1", 
           "Zone 1", "Zone 2", "Zone 2", 
           "Zone 5", "Zone 5", "Bulk"), 
  `Sort/Reporter` = c(rep("Ngn1-GFP dim (FACS sorted)", 8), rep("OMP-GFP+ (FACS sorted)", 6), "Bulk"),
  `Cell type` = c(rep("mOSN", 14), rep("Bulk", 1)),
  Replicate = c(1, 2, 3, 1, 2, 3, 1, 2, 1, 2, 1, 2, 1, 2, 1),
  check.names = FALSE
)

kable(samples, align = "l") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 11
  ) %>%
  scroll_box(width = "100%")
```

> **Note:** This table is also available as `metadata/sample_metadata.csv` in the
> course directory.

---

# 2. Connecting to your server

Connect to server via **SSH**. More instructions will follow on the day, because right now when I am writing this script I have no idea yet. 

```{bash}
ssh username@SERVER_IP
```

---

# 3. Using screen on a remote server

When running long jobs on a remote server, it is helpful to use `screen`. It keeps your
session alive even if your connection drops, and lets you run multiple things
simultaneously in separate windows.

## 3.1 Basic Screen commands

| Command | What it does |
|---------|-------------|
| `screen -S name` | Create new named session |
| `screen -ls` | List all sessions |
| `Ctrl + A, D` | Detach from current session |
| `screen -r name` | Reattach to named session |
| `screen -r` | Reattach (if only one session) |
| `screen -X -S name quit` | Kill a named session |
| `exit` | Exit and kill current session |

> **Tip:** Give your screens descriptive names (e.g. `fetchngs`, `rnaseq`,
> `chipseq`) so you can easily find them. You can have as many screens running
> simultaneously as you need.

---

# 4. Setting up the environment

## 4.1 Install Conda

We use **Miniforge3** as our conda distribution. It is lightweight, open-source,
and comes with the `conda-forge` channel set as default.

Check if conda is already installed:

```{bash}
conda --version
conda config --set auto_activate_base false # to disable (base) to be activated by default
```

If not, install Miniforge3:

```{bash}
# Download the installer
curl -L https://github.com/conda-forge/miniforge/releases/latest/download/Miniforge3-Linux-x86_64.sh -o Miniforge3.sh

# Run the installer — accept the defaults
bash Miniforge3.sh

# Reload your shell
source ~/.bashrc

# Verify installation
conda --version
```

## 4.2 Create the Nextflow environment

[Nextflow](https://www.nextflow.io/) is a workflow management system designed for scalable and reproducible scientific pipelines. It is widely used in bioinformatics - most nf-core pipelines, including `fetchngs`, are built on top of it.
Nextflow handles the complexity of running multi-step pipelines by:

- Automatically managing dependencies between steps
- Running independent steps in parallel
- Submitting jobs to computing infrastructure (local, HPC cluster, or cloud)
- Using containers (Docker, Singularity/Apptainer) to ensure reproducibility — every tool runs in an isolated, version-controlled software environment

> **Question:** How is this different from running a bash script that calls each
> tool sequentially?

Also Nextflow has a dynamic community called **nf-core** that develops and maintains a curated collection of Nextflow pipelines following best practices. Some of them we will use in this course:

| Pipeline | Purpose |
|----------|------|
| [nf-co.re/fetchngs](https://nf-co.re/fetchngs) | Download raw sequencing data from public repositories |
| [nf-co.re/rnaseq](https://nf-co.re/rnaseq/) | RNA-seq data processing (we won't use it in this course, but it exists, you know) |
| [nf-co.re/chipseq](https://snakemake.readthedocs.io) | ChIP-seq data processing |

We create a dedicated conda environment for Nextflow and Python, isolating it from other tools:

```{bash}
# Check if nextflow is already available:
nextflow -version

# If not, install it:
conda create -n nextflow-env -c bioconda -c conda-forge nextflow openjdk=17

# Check if Python 3.9 is already available:
python --version

# If not, install it:
conda create -n py39 python=3.9 -y # needed for some background processes

```

> **Note:** We explicitly install `openjdk=17` because the `nf-schema` plugin
> used by nf-core pipelines requires Java 17+. Without this, you may encounter
> errors about incompatible Java class file versions.

---

# 5. Downloading the RNA-seq and ChIP-seq data

We use [nf-core/fetchngs](https://nf-co.re/fetchngs) to download raw sequencing
data directly from SRA using accession numbers.

- Read the pipeline documentation
- Understand the structure of the pipeline and input requirements

## 5.1 Configure the input for `fetchngs`

Navigate to `~/multimodal-ngs-ubds-2026/pipelines/fetchngs`. Here you will find the following things:

- a directory `solutions` that will be our mental support in case we get stuck
- a file `nextflow.config` where I tried to put together the resources to be used by the pipeline.

Execute the following commands:
```{bash}
mkdir output # we will use this directory for the output of fetchngs
mkdir logs # for nextflow log files
touch ids.csv # we need to specify our samples with SRR accession numbers in this csv file 
touch run_fetchngs.sh # we will use this script to run fetchngs
chmod +x run_fetchngs.sh # make the .sh file executable

```

Add to `ids.csv` the SRR accession numbers from the table above.

```{bash}
cat > ids.csv << EOF
SRR12761392 
SRR12761393
SRR12761394
SRR12761403
SRR12761404
SRR12761405
SRR12761412
SRR12761413
SRR12761446 
SRR12761451
SRR12761448
SRR12761452
SRR12761450
SRR12761453
SRR12761463
EOF
```

> **Note:** You can use [nano](https://www.nano-editor.org/), [vim](https://www.vim.org/) or [neovim](https://neovim.io/) for editing the text files in console. 
> 

Prepare the previously created `run_fetchngs.sh` file by adding there a command for running `fetchngs` pipeline:

```{bash}
#!/bin/bash

export NXF_VER=24.10.6

nextflow run nf-core/fetchngs \
    --input ids.csv \
    --outdir ./output \
    --download_method ftp \
    -c nextflow.config \
    -resume
```

Downloading the data can take some time. Therefore, it could be helpful to run it inside a screen session:

```{bash}
# Navigate to your fetchngs directory
cd ~/multimodal-ngs-ubds-2026/pipelines/fetchngs

# Open a new screen
screen -S fetchngs

# Activaate the conda environments needed for the pipeline execution.
conda activate py39
conda activate --stack nextflow-env
```

## 5.2 Run `fetchngs`

The checkpoints to go through before launcing the pipeline:

- Make sure your input is correct
- Make sure you have defined the output directories
- Make sure you have your `nextflow-env` and `py39` environment active
- Start the pipeline

```{bash}
# Run the script
./run_fetchngs.sh

# Detach — the download will continue in the background
# Ctrl + A, then D

# Come back later to check progress
screen -r fetchngs
tail -f logs/fetchngs.log
```

---

# 6. Renaming the downloaded files

After `fetchngs` completes, files will have SRR-based names. We rename them to
meaningful sample names that reflect technique, zone, cell type, and replicate number. 
The names to be renamed with can be found in `sample_metadata.csv` column `Renamed to` or in this tutorial file in section `1.2 Data on NCBI GEO`

Execute the following commands in your `/fetchngs` directory:

```{bash}
mkdir fastq_renamed # we will use this directory to save here the renamed fastq files
touch rename.sh # we will use this script to rename our fastq files to a readable format
chmod +x rename.sh # make the .sh file executable
```

Modify your `rename.sh` file by including there a script for renaming the files. Make sure you double check it with the table `sample_metadata.csv` or section `1.2 Data on NCBI GEO`

```{bash, echo=TRUE, include=TRUE}
#!/bin/bash

set -euo pipefail

FASTQ_DIR="./output/fastq"  # fetchngs output directory
RENAMED_DIR="./fastq_renamed"  # destination for renamed files

mkdir -p "${RENAMED_DIR}"

# SRR accession → custom sample name
declare -A SAMPLES=(
    ["SRR12761392"]="RNA_Zone1_mOSN_Rep1"
    ["SRR12761393"]="RNA_Zone1_mOSN_Rep2"
    ["SRR12761394"]="RNA_Zone1_mOSN_Rep3"
    ["SRR12761403"]="RNA_Zone5_mOSN_Rep1"
    ["SRR12761404"]="RNA_Zone5_mOSN_Rep2"
    ["SRR12761405"]="RNA_Zone5_mOSN_Rep3"
    ["SRR12761412"]="RNA_Zone2_mOSN_Rep1"
    ["SRR12761413"]="RNA_Zone2_mOSN_Rep2"
    ["SRR12761446"]="ChIP_Zone1_mOSN_Rep1"
    ["SRR12761451"]="ChIP_Zone1_mOSN_Rep2"
    ["SRR12761448"]="ChIP_Zone2_mOSN_Rep1"
    ["SRR12761452"]="ChIP_Zone2_mOSN_Rep2"
    ["SRR12761450"]="ChIP_Zone5_mOSN_Rep1"
    ["SRR12761453"]="ChIP_Zone5_mOSN_Rep2"
    ["SRR12761463"]="ChIP_Bulk_mOSN_Rep1"
)

for SRR in "${!SAMPLES[@]}"; do
    NEW_NAME="${SAMPLES[$SRR]}"

    R1_SRC=$(ls "${FASTQ_DIR}/"*"${SRR}_1.fastq.gz" 2>/dev/null | head -1)
    R2_SRC=$(ls "${FASTQ_DIR}/"*"${SRR}_2.fastq.gz" 2>/dev/null | head -1)

    if [[ -f "${R1_SRC}" && -f "${R2_SRC}" ]]; then
        cp "${R1_SRC}" "${RENAMED_DIR}/${NEW_NAME}_R1.fastq.gz"
        cp "${R2_SRC}" "${RENAMED_DIR}/${NEW_NAME}_R2.fastq.gz"
        echo "[OK] ${SRR} → ${NEW_NAME}_R1/R2.fastq.gz"
    else
        echo "[WARN] Files not found for ${SRR}, skipping."
    fi
done

echo ""
echo "Done. Renamed files are in: ${RENAMED_DIR}"

```

Run `rename.sh` script.
```{bash}
./rename.sh 
```

You should see these lines appearing on your terminal screen as the execution of the script progresses:
```{bash}
[OK] SRR12761448 → ChIP_Zone2_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761446 → ChIP_Zone1_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761412 → RNA_Zone2_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761413 → RNA_Zone2_mOSN_Rep2_R1/R2.fastq.gz
[OK] SRR12761463 → ChIP_Bulk_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761403 → RNA_Zone5_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761405 → RNA_Zone5_mOSN_Rep3_R1/R2.fastq.gz
[OK] SRR12761404 → RNA_Zone5_mOSN_Rep2_R1/R2.fastq.gz
[OK] SRR12761393 → RNA_Zone1_mOSN_Rep2_R1/R2.fastq.gz
[OK] SRR12761392 → RNA_Zone1_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761394 → RNA_Zone1_mOSN_Rep3_R1/R2.fastq.gz
[OK] SRR12761452 → ChIP_Zone2_mOSN_Rep2_R1/R2.fastq.gz
[OK] SRR12761453 → ChIP_Zone5_mOSN_Rep2_R1/R2.fastq.gz
[OK] SRR12761450 → ChIP_Zone5_mOSN_Rep1_R1/R2.fastq.gz
[OK] SRR12761451 → ChIP_Zone1_mOSN_Rep2_R1/R2.fastq.gz
```

Re-confirm that the names match our metadata table. Navigate to `/fastq_renamed` directory and check the output.

---

# Summary of today

| Step | Tool | Output |
|------|------|--------|
| Data discovery | NCBI GEO | SRR accession numbers |
| Data download | nf-core/fetchngs | Raw FASTQ files |
| Sample renaming | bash script | Renamed FASTQ files |

---

*By Lisa Korneeva with some help of Claude Sonnet 4.6*
