---
title: "Multimodal NGS analysis of gene expression and chromatin state in olfactory epithelium"
subtitle: "Day 3 — Exploring 3t-seq results"
author: "Lisa Korneeva"
date: today
format:
  html:
    grid:
      body-width: 1000px
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    theme:
      - flatly
      - custom.scss
    highlight-style: github
    code-fold: show
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

---

# Overview

Welcome to Day 3! Yesterday we configured and launched the 3t-seq pipeline.
Today we will explore the results it produced:

- Generate and explore the **interactive report** produced by datavzrd
- Inspect the **MultiQC report** to assess data quality
- Explore interactive **MA and volcano plots** from `report.html`
- Load and explore the **DESeq2 results** in R

By the end of the day you should have a clear picture of which genes are
differentially expressed between olfactory zones and what the data quality looks like.

---

# 1. Pipeline output exploration

Let's generate the 3t-seq report. Navigate to 3t-seq directory:

```{bash}

cd ~/multimodal-ngs-ubds-2026/pipelines/rna-seq-zone-specific-3t-seq/3t-seq

```

Modify the `run.sh` file by adding `--report report-$(date -I).zip`. The final command should look like this:

```{bash}
pixi run -e dev snakemake \
    --profile profile/slurm-ubds \
    --configfile config/config.yaml \
    --rerun-incomplete \
    --report report-$(date -I).zip
```

The key output directories are:

| Directory | Contents |
|-----------|----------|
|`report-$(date -I).zip` | Report from datavzrd |
| `results/qc/multiqc/` | MultiQC quality control reports |
| `results/analysis/rdata/deseq2/` | DESeq2 R objects (RDS) |
| `results/alignments/*.ReadsPerGene.out.tab` | Gene counts |

---

# 2. Quality Control — MultiQC Report

## 2.1 What is MultiQC?

[MultiQC](https://multiqc.info) aggregates quality control metrics from multiple
tools and samples into a single interactive HTML report. The 3t-seq pipeline
runs MultiQC after each major step, collecting metrics from FastQC, Trimmomatic,
STAR, featureCounts, and more.

## 2.2 Opening the MultiQC Report

Open the MultiQC report for one of the comparisons in your browser:

```{bash}
# Path to MultiQC report
~/results/qc/multiqc/zone1_vs_zone2/multiqc_report.html
```

## 2.3 Interpreting the MultiQC Report

Work through the following sections of the report and answer the questions below.

### General Statistics

The General Statistics table at the top gives a high-level overview of each sample.

**Exercise:** Look at the General Statistics table and answer:

1. How many reads were sequenced per sample?
2. What percentage of reads passed quality filtering?
3. Are there any samples that look notably different from the others?

### FastQC — Sequence Quality

FastQC assesses the quality of raw and trimmed reads.

**Exercise:**

1. Look at the **Per Base Sequence Quality** plot. What does it show?
   At what position does quality start to drop?
2. Look at the **Per Sequence GC Content** plot. Does the distribution
   look normal? What would an abnormal distribution suggest?
3. Compare the raw vs trimmed reads sections. Did trimming improve the quality?

### STAR Alignment Statistics

**Exercise:**

1. What percentage of reads aligned uniquely to the genome?
2. What percentage of reads were multi-mapped?
3. Are alignment rates consistent across samples?

> **Note:** For RNA-seq, a uniquely mapped rate of >70% is generally considered
> acceptable. Rates below this may indicate contamination, wrong reference genome,
> or library quality issues.

### featureCounts — Gene Assignment

**Exercise:**

1. What percentage of reads were successfully assigned to genes?
2. What are the main reasons for reads not being assigned?

---

# 3. DESeq2 results

## 3.1 What did DESeq2 do?

The 3t-seq pipeline ran [DESeq2](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) — a statistical method for differential
expression analysis — to compare gene expression between zones. DESeq2:

1. Takes raw read counts per gene per sample
2. Normalises for sequencing depth differences between samples
3. Models count data using a negative binomial distribution
4. Tests for differential expression between conditions
5. Applies multiple testing correction (Benjamini-Hochberg)

The results tell us which genes are significantly up- or down-regulated between
zones.

Explore the MA plots and volcano plots from the `report.html` generated by 3t-seq. 
- What is the trend of gene expression changes in our comparisons?
- Which family do the significantly regulated genes belong to?
- How does the gene expression changes within comparison if you increase/decrease the thresholds for the Log2FoldChange / p-value?

---

# 4. Exploring the data in R



## 4.1 PCA Plot

A **Principal Component Analysis (PCA)** plot shows how similar samples are to
each other overall. Samples from the same condition should cluster together.

```{r}
# Variance stabilising transformation for visualisation
vsd <- vst(dds, blind = FALSE)

# PCA plot
plotPCA(vsd, intgroup = "zone") +
  theme_bw() +
  ggtitle("PCA — Zone 1 vs Zone 2") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Exercise:**

1. Do samples cluster by zone as expected?
2. Is there any sample that looks like an outlier?
3. What percentage of variance is explained by PC1 and PC2?

## 4.2 Sample-to-Sample Distance Heatmap

```{r}
library(pheatmap)

# Calculate sample distances
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)

# Plot heatmap
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         main = "Sample-to-sample distances")
```

**Exercise:** Which samples are most similar to each other?

---

## 3.2 Set Up Your R Environment

```{r}
# Install packages if needed
# install.packages(c("ggplot2", "ggrepel", "dplyr", "readr"),
#                  repos = "https://packagemanager.posit.co/cran/__linux__/jammy/latest")
# BiocManager::install("DESeq2")

library(DESeq2)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(readr)
```

## 3.3 Define Paths

```{r}
# Set the base path to your results directory
results_dir <- "/path/to/pipelines/rnaseq/3t-seq/results"

# Define paths for each comparison
comparisons <- c("zone1_vs_zone2", "zone1_vs_zone5", "zone2_vs_zone5")

# Path to DESeq2 results tables
tables_dir <- file.path(results_dir, "analysis", "tables", "deseq2")

# Path to DESeq2 RDS objects
rdata_dir <- file.path(results_dir, "analysis", "rdata", "deseq2")
```

## 3.4 Load the DESeq2 Object

```{r}
# Load the DESeq2 object for zone1 vs zone2 comparison
dds <- readRDS(file.path(rdata_dir, "zone1_vs_zone2", "dds.rds"))

# Inspect the object
dds
dim(dds)         # genes x samples
colData(dds)     # sample metadata
```

**Exercise:** Inspect the DESeq2 object:

1. How many genes are in the dataset?
2. How many samples?
3. What does `colData(dds)` show you?

## 3.5 Load the Results Tables

The pipeline produces two results tables per comparison:

- **`results.csv`** — raw DESeq2 results with standard log2 fold changes
- **`results.shrink.csv`** — results with shrunken log2 fold changes (recommended for visualisation)

```{r}
# Load raw results
res <- read_csv(file.path(tables_dir, "zone1_vs_zone2", "results.csv"))

# Load shrinkage results
res_shrink <- read_csv(file.path(tables_dir, "zone1_vs_zone2", "results.shrink.csv"))

# Inspect
head(res)
colnames(res)
dim(res)
```

The key columns in the results table are:

| Column | Description |
|--------|-------------|
| `gene_id` | Ensembl gene ID |
| `gene_name` | Gene symbol |
| `baseMean` | Average normalised count across all samples |
| `log2FoldChange` | Log2 fold change (positive = higher in zone1) |
| `lfcSE` | Standard error of the log2 fold change |
| `pvalue` | Raw p-value |
| `padj` | Adjusted p-value (Benjamini-Hochberg) |

**Exercise:**

1. How many genes have an adjusted p-value below 0.05?
2. How many of these are upregulated in Zone 1 vs Zone 2?
3. How many are downregulated?

```{r}
# Your code here
# Hint: use filter() from dplyr

```

---

# 5. Volcano Plot

A **volcano plot** visualises differential expression results by plotting:

- X-axis: log2 fold change (effect size)
- Y-axis: -log10 adjusted p-value (significance)

Genes in the upper left/right corners are both significant AND strongly changed —
these are the most biologically interesting.

## 5.1 Basic Volcano Plot

```{r}
# Prepare data — remove NA values
res_clean <- res %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange > 1  ~ "Up in Zone 1",
      padj < 0.05 & log2FoldChange < -1 ~ "Up in Zone 2",
      TRUE                               ~ "Not significant"
    )
  )

# Count genes per category
table(res_clean$significance)
```

```{r}
# Basic volcano plot
ggplot(res_clean, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_manual(values = c(
    "Up in Zone 1"     = "#E41A1C",
    "Up in Zone 2"     = "#377EB8",
    "Not significant"  = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  theme_bw() +
  labs(
    title    = "Zone 1 vs Zone 2",
    x        = "Log2 fold change",
    y        = "-Log10 adjusted p-value",
    colour   = NULL
  ) +
  theme(
    plot.title  = element_text(hjust = 0.5),
    legend.position = "top"
  )
```

## 5.2 Volcano Plot with Gene Labels

Label the top differentially expressed genes:

```{r}
# Select top genes to label
top_genes <- res_clean %>%
  filter(significance != "Not significant") %>%
  arrange(padj) %>%
  head(20)

# Volcano plot with labels
ggplot(res_clean, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_text_repel(
    data          = top_genes,
    aes(label     = gene_name),
    size          = 3,
    max.overlaps  = 20,
    colour        = "black"
  ) +
  scale_colour_manual(values = c(
    "Up in Zone 1"    = "#E41A1C",
    "Up in Zone 2"    = "#377EB8",
    "Not significant" = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  theme_bw() +
  labs(
    title  = "Zone 1 vs Zone 2",
    x      = "Log2 fold change",
    y      = "-Log10 adjusted p-value",
    colour = NULL
  ) +
  theme(
    plot.title      = element_text(hjust = 0.5),
    legend.position = "top"
  )
```

**Exercise:**

1. How many genes are significantly upregulated in Zone 1 (padj < 0.05, log2FC > 1)?
2. How many are significantly upregulated in Zone 2?
3. Do you recognise any of the top gene names? Are any of them olfactory receptors?
4. Repeat the volcano plot for the zone1_vs_zone5 and zone2_vs_zone5 comparisons.
   Do the same genes appear consistently?

> **Solution:** Will be available by the end of the day

```{r, echo=FALSE, include=FALSE}
# Solution: function to make volcano plot for any comparison
make_volcano <- function(comparison, lfc_threshold = 1, padj_threshold = 0.05) {
  
  res <- read_csv(file.path(tables_dir, comparison, "results.csv"))
  
  zones <- strsplit(comparison, "_vs_")[[1]]
  zone1 <- gsub("zone", "Zone ", zones[1])
  zone2 <- gsub("zone", "Zone ", zones[2])
  
  res_clean <- res %>%
    filter(!is.na(padj), !is.na(log2FoldChange)) %>%
    mutate(
      significance = case_when(
        padj < padj_threshold & log2FoldChange >  lfc_threshold ~ paste("Up in", zone1),
        padj < padj_threshold & log2FoldChange < -lfc_threshold ~ paste("Up in", zone2),
        TRUE ~ "Not significant"
      )
    )
  
  top_genes <- res_clean %>%
    filter(significance != "Not significant") %>%
    arrange(padj) %>%
    head(20)
  
  ggplot(res_clean, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
    geom_point(alpha = 0.5, size = 0.8) +
    geom_text_repel(data = top_genes, aes(label = gene_name),
                    size = 3, max.overlaps = 20, colour = "black") +
    scale_colour_manual(values = c(
      paste("Up in", zone1) = "#E41A1C",
      paste("Up in", zone2) = "#377EB8",
      "Not significant"     = "grey70"
    )) +
    geom_vline(xintercept = c(-lfc_threshold, lfc_threshold),
               linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(padj_threshold),
               linetype = "dashed", colour = "grey40") +
    theme_bw() +
    labs(title  = gsub("_", " ", comparison),
         x      = "Log2 fold change",
         y      = "-Log10 adjusted p-value",
         colour = NULL) +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "top")
}

# Make plots for all comparisons
for (comp in comparisons) {
  print(make_volcano(comp))
}
```

---

# 6. Top Differentially Expressed Genes

## 6.1 Expression Heatmap of Top DE Genes

```{r}
# Get top 50 DE genes by adjusted p-value
top50 <- res_clean %>%
  filter(significance != "Not significant") %>%
  arrange(padj) %>%
  head(50) %>%
  pull(gene_id)

# Extract normalised counts for these genes
norm_counts <- counts(dds, normalized = TRUE)
top50_counts <- norm_counts[rownames(norm_counts) %in% top50, ]

# Add gene names as row labels
gene_names <- res_clean %>%
  filter(gene_id %in% top50) %>%
  select(gene_id, gene_name) %>%
  distinct()

rownames(top50_counts) <- gene_names$gene_name[match(rownames(top50_counts), gene_names$gene_id)]

# Plot heatmap
pheatmap(log2(top50_counts + 1),
         scale          = "row",
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         fontsize_row   = 7,
         main           = "Top 50 DE genes — Zone 1 vs Zone 2",
         annotation_col = as.data.frame(colData(dds)[, "zone", drop = FALSE]))
```

## 6.2 Individual Gene Expression

Plot the expression of a specific gene across samples:

```{r}
# Plot expression of a specific gene
# Replace "Olfr1507" with any gene you are interested in
gene_of_interest <- "Olfr1507"

gene_id <- res_clean %>%
  filter(gene_name == gene_of_interest) %>%
  pull(gene_id)

plotCounts(dds,
           gene      = gene_id,
           intgroup  = "zone",
           returnData = TRUE) %>%
  ggplot(aes(x = zone, y = count, colour = zone)) +
  geom_jitter(width = 0.1, size = 3) +
  scale_y_log10() +
  theme_bw() +
  labs(title = gene_of_interest,
       x     = "Zone",
       y     = "Normalised count (log10)") +
  theme(plot.title      = element_text(hjust = 0.5),
        legend.position = "none")
```

**Exercise:** Pick 3 genes from your volcano plot that interest you and plot their
expression. Are they olfactory receptors? Do they make biological sense given
what you know about zone-specific expression in the MOE?

---

# 7. Interactive Results Table

The pipeline also generates an interactive results table using
[datavzrd](https://datavzrd.github.io). Open it in your browser:

```{bash}
results/analysis/datavzrd/zone1_vs_zone2/datavzrd/index.html
```

This allows you to:
- Sort and filter genes by any column
- Search for specific genes by name
- Visualise fold changes and significance interactively

**Exercise:** Use the datavzrd table to find all olfactory receptor genes
(gene names starting with "Olfr") that are significantly differentially
expressed between Zone 1 and Zone 2. How many are there?

---

# Summary of Today

| Step | Tool | What we learned |
|------|------|-----------------|
| Quality control | MultiQC | Data quality, alignment rates, gene assignment |
| DESeq2 results | R/DESeq2 | Which genes are differentially expressed |
| PCA | R/DESeq2 | Sample similarity and batch effects |
| Volcano plot | R/ggplot2 | Effect size vs significance |
| Heatmap | R/pheatmap | Expression patterns across samples |
| Interactive table | datavzrd | Exploring results interactively |

Tomorrow we will move on to the ChIP-seq data and investigate the epigenomic
landscape of the olfactory epithelium.

---

# Quick Reference

## Key thresholds for differential expression

| Threshold | Typical value | What it means |
|-----------|--------------|---------------|
| Adjusted p-value | < 0.05 | Statistically significant after multiple testing correction |
| Log2 fold change | > 1 or < -1 | At least 2-fold difference in expression |
| baseMean | > 10 | Expressed at a minimum level (filter out lowly expressed genes) |

## Useful R commands

```{r}
# Summary of DE results
summary(results(dds))

# Number of significant genes
sum(res$padj < 0.05, na.rm = TRUE)

# Top upregulated genes
res %>% filter(padj < 0.05, log2FoldChange > 0) %>% arrange(desc(log2FoldChange)) %>% head(10)

# Top downregulated genes
res %>% filter(padj < 0.05, log2FoldChange < 0) %>% arrange(log2FoldChange) %>% head(10)
```

---

*By Lisa Korneeva with some help of Claude Sonnet 4.6*
