---
title: "Multimodal NGS analysis of gene expression and chromatin state in olfactory epithelium"
subtitle: "Day 3 — RNA-seq results — Solutions"
author: "Lisa Korneeva"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    theme:
      - flatly
      - custom.scss
    highlight-style: github
    code-fold: show
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

# Overview

Today we explore the results of the RNA-seq pipeline. We will:

- Load raw count data from STAR output
- Construct a DESeq2 object and run differential expression analysis
- Assess data quality with PCA and sample distance heatmaps
- Visualise results with MA plots and volcano plots
- Perform gene ontology enrichment analysis with clusterProfiler

---

# 1. Setup

## 1.1 Load libraries

```{r}
library(here)
library(DESeq2)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(clusterProfiler)
library(org.Mm.eg.db)
library(biomaRt)
```

## 1.2 Define paths

```{r}
# Set paths using here() — all paths relative to project root
counts_dir <- here(
  "pipelines",
  "rna-seq-zone-specific-3t-seq",
  "solutions",
  "results",
  "alignments"
)

# List all ReadsPerGene files
count_files <- list.files(
  counts_dir,
  pattern     = "ReadsPerGene.out.tab",
  recursive   = TRUE,
  full.names  = TRUE
)

# Extract sample names from file paths
sample_names <- basename(count_files) |>
  str_remove("\\.ReadsPerGene\\.out\\.tab$")

names(count_files) <- sample_names
count_files
```

---

# 2. Load and prepare count data

## 2.1 Load ReadsPerGene files

STAR's `ReadsPerGene.out.tab` has 4 columns:
- Column 1: gene ID
- Column 2: unstranded counts
- Column 3: forward-stranded counts ← we use this (NuQuant kit)
- Column 4: reverse-stranded counts

The first 4 rows are summary statistics (N_unmapped, N_multimapping, etc.) — we skip these.

```{r}
# Load all count files and extract forward-stranded counts (column 3)
count_list <- lapply(count_files, function(f) {
  read.table(f, header = FALSE, skip = 4, row.names = 1)[, 2, drop = FALSE]
})

# Combine into one count matrix
count_matrix <- do.call(cbind, count_list)
colnames(count_matrix) <- names(count_files)

dim(count_matrix)   # should be 55450 genes x 8 samples
head(count_matrix)
```

## 2.2 Add gene symbols

```{r}
# Connect to Ensembl biomaRt for mouse
mart <- useEnsembl(
  biomart = "ensembl",
  dataset = "mmusculus_gene_ensembl",
  version = 105
)

gene_info <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
  filters    = "ensembl_gene_id",
  values     = rownames(count_matrix),
  mart       = mart
)

# Check if Olfr names are there
gene_info |> filter(grepl("^Olfr", external_gene_name)) |> head(10)

head(gene_info)
```

## 2.3 Create sample metadata

```{r}
# Create colData — sample metadata for DESeq2
coldata <- data.frame(
  sample    = colnames(count_matrix),
  zone      = factor(c("zone1", "zone1", "zone1",
                       "zone2", "zone2",
                       "zone5", "zone5", "zone5")),
  replicate = c(1, 2, 3, 1, 2, 1, 2, 3)
)
rownames(coldata) <- coldata$sample
coldata
```

---

# 3. DESeq2 analysis

## 3.1 Create DESeq2 object

```{r}
# Subset count matrix to samples in coldata (in case order differs)
count_matrix_ordered <- count_matrix[, rownames(coldata)]

# Create DESeqDataSet
dds_all <- DESeqDataSetFromMatrix(
  countData = count_matrix_ordered,
  colData   = coldata,
  design    = ~ zone
)

# Pre-filter: remove genes with very low counts
keep <- rowSums(counts(dds_all) >= 10) >= 3
dds_all <- dds_all[keep, ]
nrow(dds_all)
```

## 3.2 Run DESeq2

```{r}
dds_all <- DESeq(dds_all)
```

---

# 4. Quality control

## 4.1 PCA

```{r}
# Variance stabilising transformation
vsd <- vst(dds_all, blind = FALSE)

# PCA
pca_data <- plotPCA(vsd, intgroup = "zone", returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, colour = zone, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3) +
  scale_colour_manual(values = c(
    zone1 = "#E41A1C",
    zone2 = "#377EB8",
    zone5 = "#4DAF4A"
  )) +
  xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  theme_bw() +
  ggtitle("PCA — all samples") +
  theme(plot.title = element_text(hjust = 0.5))
```

## 4.2 Sample distance heatmap

```{r}
# Sample-to-sample distances
sampleDists     <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)

# Annotation for heatmap
annotation_col <- data.frame(
  zone = coldata$zone,
  row.names = rownames(coldata)
)

pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  annotation_col           = annotation_col,
  main                     = "Sample-to-sample distances"
)
```

---

# 5. Pairwise comparisons

## 5.1 Extract results for each comparison

```{r}
# Helper function to extract and annotate DESeq2 results
get_results <- function(dds, contrast, gene_info) {
  res <- results(dds, contrast = contrast, alpha = 0.05)
  res <- lfcShrink(dds, contrast = contrast, res = res, type = "ashr")
  res_df <- as.data.frame(res) |>
    rownames_to_column("ensembl_gene_id") |>
    left_join(gene_info, by = "ensembl_gene_id") |>
    arrange(padj)
  return(res_df)
}

# Zone 1 vs Zone 2
res_z1_z2 <- get_results(
  dds_all,
  contrast  = c("zone", "zone1", "zone2"),
  gene_info = gene_info
)

# Zone 1 vs Zone 5
res_z1_z5 <- get_results(
  dds_all,
  contrast  = c("zone", "zone1", "zone5"),
  gene_info = gene_info
)

# Zone 2 vs Zone 5
res_z2_z5 <- get_results(
  dds_all,
  contrast  = c("zone", "zone2", "zone5"),
  gene_info = gene_info
)

head(res_z2_z5)
```

## 5.2 Summary — how many genes are DE?

```{r}
# Count significant genes per comparison
summarise_de <- function(res_df, name) {
  sig <- res_df |> filter(!is.na(padj), padj < 0.05)
  data.frame(
    comparison   = name,
    total_sig    = nrow(sig),
    upregulated  = sum(sig$log2FoldChange > 1),
    downregulated = sum(sig$log2FoldChange < -1)
  )
}

de_summary <- bind_rows(
  summarise_de(res_z1_z2, "Zone1 vs Zone2"),
  summarise_de(res_z1_z5, "Zone1 vs Zone5"),
  summarise_de(res_z2_z5, "Zone2 vs Zone5")
)

de_summary
```

---

# 6. Visualisation

## 6.1 MA plot

An MA plot shows log2 fold change (M) vs mean expression (A). Significant genes are highlighted in red.

```{r}
plot_ma <- function(res_df, title) {
  res_df |>
    filter(!is.na(padj)) |>
    mutate(significant = padj < 0.05 & abs(log2FoldChange) > 1) |>
    ggplot(aes(x = log10(baseMean + 1),
               y = log2FoldChange,
               colour = significant)) +
    geom_point(alpha = 0.4, size = 0.8) +
    scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#E41A1C")) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_hline(yintercept = c(-1, 1), linetype = "dotted", colour = "grey40") +
    theme_bw() +
    labs(title  = title,
         x      = "Log10 mean expression",
         y      = "Log2 fold change",
         colour = "Significant") +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "top")
}

plot_ma(res_z1_z2, "MA plot — Zone 1 vs Zone 2")
plot_ma(res_z1_z5, "MA plot — Zone 1 vs Zone 5")
plot_ma(res_z2_z5, "MA plot — Zone 2 vs Zone 5")
```

## 6.2 Volcano plot

```{r}
plot_volcano <- function(res_df, title,
                         lfc_threshold  = 1,
                         padj_threshold = 0.05,
                         n_labels       = 20) {

  zones  <- strsplit(title, " vs ")[[1]]
  up_lab <- paste("Up in", zones[1])
  dn_lab <- paste("Up in", zones[2])

  res_clean <- res_df |>
    filter(!is.na(padj), !is.na(log2FoldChange)) |>
    mutate(significance = case_when(
      padj < padj_threshold & log2FoldChange >  lfc_threshold ~ up_lab,
      padj < padj_threshold & log2FoldChange < -lfc_threshold ~ dn_lab,
      TRUE ~ "Not significant"
    ))

  top_genes <- res_clean |>
    filter(significance != "Not significant") |>
    arrange(padj) |>
    head(n_labels)

  ggplot(res_clean,
         aes(x = log2FoldChange,
             y = -log10(padj),
             colour = significance)) +
    geom_point(alpha = 0.5, size = 0.8) +
    geom_text_repel(
      data         = top_genes,
      aes(label    = external_gene_name),
      size         = 3,
      max.overlaps = 20,
      colour       = "black"
    ) +
    scale_colour_manual(values = c(
      setNames("#E41A1C", up_lab),
      setNames("#377EB8", dn_lab),
      "Not significant" = "grey70"
    )) +
    geom_vline(xintercept = c(-lfc_threshold, lfc_threshold),
               linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(padj_threshold),
               linetype = "dashed", colour = "grey40") +
    theme_bw() +
    labs(title  = title,
         x      = "Log2 fold change",
         y      = "-Log10 adjusted p-value",
         colour = NULL) +
    theme(plot.title      = element_text(hjust = 0.5),
          legend.position = "top")
}

plot_volcano(res_z1_z2, "Zone 1 vs Zone 2")
plot_volcano(res_z1_z5, "Zone 1 vs Zone 5")
plot_volcano(res_z2_z5, "Zone 2 vs Zone 5")
```

---

# 7. Gene ontology enrichment analysis

```{r}
# Get significantly upregulated genes in Zone 1 vs Zone 2
up_z1 <- res_z1_z2 |>
  filter(!is.na(padj), padj < 0.05, log2FoldChange > 1) |>
  pull(ensembl_gene_id)

# Run GO enrichment
ego_z1 <- enrichGO(
  gene          = up_z1,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",           # Biological Process
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# Dotplot
dotplot(ego_z1, showCategory = 20, title = "GO Biological Process — Zone 1 upregulated")
```

```{r}
# Barplot
barplot(ego_z1, showCategory = 15, title = "GO Biological Process — Zone 1 upregulated")
```

```{r}
# Export results table
ego_df <- as.data.frame(ego_z1)
head(ego_df[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")])
```

---

# 8. Save results

```{r}
# Create output directory if it doesn't exist
dir.create(here("analysis", "solutions"), recursive = TRUE, showWarnings = FALSE)

# Save DESeq2 results tables
write_csv(res_z1_z2, here("analysis", "solutions", "results_zone1_vs_zone2.csv"))
write_csv(res_z1_z5, here("analysis", "solutions", "results_zone1_vs_zone5.csv"))
write_csv(res_z2_z5, here("analysis", "solutions", "results_zone2_vs_zone5.csv"))

# Save DESeq2 object — contains normalised counts, size factors, dispersion
# estimates and design for all 8 samples across all three zones
saveRDS(dds_all, here("analysis", "solutions", "dds_all.rds"))

# Save gene annotation — needed for joining gene symbols on Day 5
saveRDS(gene_info, here("analysis", "solutions", "gene_info.rds"))
```

---

*By Lisa Korneeva with some help of Claude Sonnet 4.6*
