---
title: "Multimodal NGS analysis of gene expression and chromatin state in olfactory epithelium"
subtitle: "Day 3 — RNA-seq results — Exercises"
author: "Lisa Korneeva"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    theme:
      - flatly
      - custom.scss
    highlight-style: github
    code-fold: show
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

# Overview

Today we explore the results of the RNA-seq pipeline. We will:

- Load raw count data from STAR output
- Construct a DESeq2 object and run differential expression analysis
- Assess data quality with PCA and sample distance heatmaps
- Visualise results with MA plots and volcano plots
- Perform gene ontology enrichment analysis with clusterProfiler

---

# 1. Setup

## 1.1 Load libraries

Load the following libraries. Install any that are missing using `BiocManager::install()` or `install.packages()`.

```{r}
library(here)
library(DESeq2)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(clusterProfiler)
library(org.Mm.eg.db)
library(biomaRt)
```

## 1.2 Define paths

Use `here()` to define the path to your count files. The `ReadsPerGene.out.tab` files are located in the `/home/YOUR_USER/multimodal-ngs-course-2026/pipelines/rna-seq-zone-specific-3t-seq/solutions/results/alignments` directory of the pipeline output.

```{r}
# Define path to count files using here()
counts_dir <- here(...)

# Use list.files() to find all ReadsPerGene.out.tab files recursively
count_files <- list.files(
  counts_dir,
  pattern    = "ReadsPerGene.out.tab",
  recursive  = TRUE,
  full.names = TRUE
)

# Extract sample names from paths and set as names
# Hint: use basename() on count_files
sample_names <- ...
names(count_files) <- sample_names
```

---

# 2. Load and prepare count data

## 2.1 Load ReadsPerGene files

STAR's `ReadsPerGene.out.tab` has 4 columns:

| Column | Content |
|--------|---------|
| 1 | Gene ID |
| 2 | Unstranded counts |
| 3 | Forward-stranded counts |
| 4 | Reverse-stranded counts |

> **Question:** Which column should we use for the NuQuant library preparation kit, and why?

The first 4 rows are summary statistics — skip them with `skip = 4`.

```{r}
# Load all count files using lapply()
# Read each file with read.table(), skip the first 4 rows, use column 1 as row names
# Extract only the forward-stranded column (column 2 after skipping col 1 with row.names)
count_list <- lapply(count_files, function(f) {
  # Hint: read.table(f, header = FALSE, skip = 4, row.names = 1)[, COLUMN, drop = FALSE]
  ...
})

# Combine into a count matrix using do.call(cbind, ...)
count_matrix <- ...
colnames(count_matrix) <- names(count_files)

# Check dimensions 
dim(count_matrix)
head(count_matrix)
```

## 2.2 Add gene symbols

Use `biomaRt` to retrieve gene symbols for all Ensembl IDs in your count matrix.

```{r}
# Connect to Ensembl for mouse
mart <- useEnsembl(
  biomart = "ensembl",
  dataset = "mmusculus_gene_ensembl",
  version = 105
)

gene_info <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
  filters    = "ensembl_gene_id",
  values     = rownames(count_matrix),
  mart       = mart
)

# Check if Olfr names are there
gene_info |> filter(grepl("^Olfr", external_gene_name)) |> head(10)

head(gene_info)
```

## 2.3 Create sample metadata

```{r}
# Create a data frame describing your samples
# Columns: sample, zone, replicate
# Make zone a factor
coldata <- data.frame(
  sample    = colnames(count_matrix),
  zone      = factor(c(...)),  # fill in zone assignments
  replicate = c(...)           # fill in replicate numbers
)
rownames(coldata) <- coldata$sample
coldata
```

> **Question:** Why do we need to make `zone` a factor? What happens if we leave it as a character vector?

---

# 3. DESeq2 analysis

## 3.1 Create DESeq2 object

```{r}
# Create DESeqDataSet from count matrix
# design = ~ zone (we want to test for differences between zones)
dds_all <- DESeqDataSetFromMatrix(
  countData = ...,
  colData   = ...,
  design    = ...
)

# Pre-filter: keep only genes with at least 10 counts in at least 3 samples
# Hint: rowSums(counts(dds_all) >= 10) >= 3
keep <- ...
dds_all <- dds_all[keep, ]

# How many genes remain after filtering?
nrow(dds_all)
```

> **Question:** Why do we filter out lowly expressed genes before running DESeq2?

## 3.2 Run DESeq2

```{r}
# Run the DESeq2 pipeline
# This normalises counts, estimates dispersions, and fits the negative binomial model
dds_all <- DESeq(dds_all)
```

---

# 4. Quality control

## 4.1 PCA

A PCA plot shows how similar samples are to each other overall. Samples from the same zone should cluster together.

```{r}
# Apply variance stabilising transformation (VST) for visualisation
# Hint: vst(dds_all, blind = FALSE)
vsd <- ...

# Generate PCA data
# Hint: plotPCA(vsd, intgroup = "zone", returnData = TRUE)
pca_data    <- ...
percent_var <- round(100 * attr(pca_data, "percentVar"))

# Plot PCA using ggplot2
# Colour by zone, label by sample name using geom_text_repel()
ggplot(pca_data, aes(x = PC1, y = PC2, colour = zone, label = name)) +
  ...
```

> **Questions:**
>
> 1. Do samples cluster by zone as expected?
> 2. How much variance is explained by PC1 and PC2?
> 3. Is there any sample that looks like an outlier?

## 4.2 Sample distance heatmap

```{r}
# Calculate Euclidean distances between samples in VST space
sampleDists      <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)

# Plot heatmap using pheatmap()
# Add zone annotation to columns
annotation_col <- data.frame(
  zone = coldata$zone,
  row.names = rownames(coldata)
)

pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  annotation_col           = annotation_col,
  main                     = "Sample-to-sample distances"
)
```

> **Question:** Which samples are most similar to each other? Does this make biological sense?

---

# 5. Pairwise comparisons

## 5.1 Extract DESeq2 results

```{r}
# Extract results for Zone 1 vs Zone 2
# Hint: results(dds_all, contrast = c("zone", "zone1", "zone2"), alpha = 0.05)
# Apply LFC shrinkage with lfcShrink(..., type = "apeglm")
res_z1_z2 <- ...

# Convert to data frame, add gene symbols from gene_info
# Hint: as.data.frame(res) |> rownames_to_column("ensembl_gene_id") |> left_join(gene_info, ...)
res_z1_z2 <- ...

# Repeat for Zone 1 vs Zone 5
res_z1_z5 <- ...

# Repeat for Zone 2 vs Zone 5
res_z2_z5 <- ...

head(res_z1_z2)
```

> **Question:** What does LFC shrinkage do and why is it recommended for visualisation?

## 5.2 How many genes are differentially expressed?

```{r}
# For each comparison, count:
# - Total significant genes (padj < 0.05)
# - Upregulated genes (padj < 0.05 AND log2FoldChange > 1)
# - Downregulated genes (padj < 0.05 AND log2FoldChange < -1)

# Hint: use filter() and summarise() from dplyr

# Zone 1 vs Zone 2
res_z1_z2 |>
  filter(!is.na(padj), padj < 0.05) |>
  summarise(
    total         = n(),
    upregulated   = sum(log2FoldChange > 1),
    downregulated = sum(log2FoldChange < -1)
  )
```

> **Exercise:** Do the same for Zone 1 vs Zone 5 and Zone 2 vs Zone 5.

---

# 6. Visualisation

## 6.1 MA plot

An MA plot shows:
- **X-axis:** mean expression across all samples (log scale)
- **Y-axis:** log2 fold change between conditions

Significant genes (padj < 0.05, |log2FC| > 1) should be highlighted.

```{r}
# Create an MA plot for Zone 1 vs Zone 2
# Hint: use ggplot() with geom_point()
# Colour points by significance (padj < 0.05 & abs(log2FoldChange) > 1)
# Add horizontal lines at y = 0 and y = ±1 with geom_hline()

res_z1_z2 |>
  filter(!is.na(padj)) |>
  mutate(significant = ...) |>
  ggplot(aes(x = ..., y = ..., colour = ...)) +
  geom_point(...) +
  ...
```

> **Exercise:** Create MA plots for the other two comparisons.
>
> **Question:** What do genes in the upper right of the MA plot represent?

## 6.2 Volcano plot

A volcano plot shows:

- **X-axis:** log2 fold change (effect size)
- **Y-axis:** -log10 adjusted p-value (significance)

```{r}
# Create a volcano plot for Zone 1 vs Zone 2
# Colour points by significance category:
#   - "Up in Zone 1": padj < 0.05 AND log2FoldChange > 1
#   - "Up in Zone 2": padj < 0.05 AND log2FoldChange < -1
#   - "Not significant": everything else
# Label the top 20 genes by adjusted p-value using geom_text_repel()

res_z1_z2 |>
  filter(!is.na(padj), !is.na(log2FoldChange)) |>
  mutate(significance = case_when(
    padj < 0.05 & log2FoldChange >  1 ~ "Up in Zone 1",
    padj < 0.05 & log2FoldChange < -1 ~ "Up in Zone 2",
    TRUE ~ "Not significant"
  )) |>
  ggplot(aes(x = log2FoldChange,
             y = -log10(padj),
             colour = significance)) +
  ...
```

> **Exercise:** Repeat for the other two comparisons.
>
> **Question:** Are any olfactory receptor genes (starting with "Olfr") among the top hits?
> Hint: `filter(str_starts(external_gene_name, "Olfr"))`

---

# 7. Gene ontology enrichment analysis

Gene ontology (GO) enrichment analysis tests whether a set of genes (e.g. the genes upregulated in Zone 1) is statistically enriched for specific biological functions, processes, or cellular components.

```{r}
# Get significantly upregulated genes in Zone 1 vs Zone 2
# Hint: filter(padj < 0.05, log2FoldChange > 1) |> pull(ensembl_gene_id)
up_z1 <- ...

# Run GO enrichment using enrichGO() from clusterProfiler
# OrgDb = org.Mm.eg.db (mouse annotation database)
# keyType = "ENSEMBL"
# ont = "BP" (Biological Process)
ego_z1 <- enrichGO(
  gene          = up_z1,
  OrgDb         = ...,
  keyType       = ...,
  ont           = ...,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  readable      = TRUE
)

# Visualise results
# Hint: dotplot(ego_z1, showCategory = 20)
dotplot(...)
```

> **Questions:**
>
> 1. What biological processes are enriched in Zone 1 upregulated genes?
> 2. Does this make biological sense given what you know about olfactory receptor expression?
> 3. Try running the same analysis for Zone 5 upregulated genes. Are the enriched terms different?

---

# 8. Save your results

```{r}
# Save your DESeq2 results tables as CSV files
# Hint: write_csv(res_z1_z2, here("analysis", "exercises", "results_zone1_vs_zone2.csv"))
write_csv(res_z1_z2, ...)
write_csv(res_z1_z5, ...)
write_csv(res_z2_z5, ...)

# Save the DESeq2 object for use on Day 5
saveRDS(dds_all, here("analysis", "exercises", "dds_all.rds"))
```

---

*By Lisa Korneeva with some help of Claude Sonnet 4.6*
