---
title: "Multimodal NGS analysis of gene expression and chromatin state in olfactory epithelium"
subtitle: "Day 5 - Integrative analysis of RNA-seq and ChIP-seq data - VIBE CODING WARNING"
author: "Lisa Korneeva"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    theme:
      - flatly
      - custom.scss
    highlight-style: github
    code-fold: show
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

Welcome to Day 5 - the final day of the course! Today we bring together the two
datasets we have analysed:

- **RNA-seq**: zone-specific gene expression in mOSN
- **ChIP-seq**: H3K9me3 signal across olfactory receptor loci

The central biological question is: **does H3K9me3 enrichment correlate with
silencing of olfactory receptor genes across zones?**

We expect that in a given zone, the olfactory receptor genes that are *not*
expressed should be marked by H3K9me3 (silenced heterochromatin), while the
expressed OR genes should have lower H3K9me3 signal.

---

# 1. Setup

## 1.1 Load libraries

```{r}
library(here)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
```

## 1.2 Define paths

```{r}
# ChIP-seq results directory
chipseq_dir <- here(
  "pipelines",
  "chip-seq-zone-specific-nextflow",
  "solutions"
)

# RNA-seq results directory
rnaseq_dir <- here(
  "pipelines",
  "rna-seq-zone-specific-3t-seq",
  "solutions"
)
```

---

# 2. Load the data

## 2.1 Load H3K9me3 signal

The H3K9me3 signal table was computed with deepTools `multiBigwigSummary` over
olfactory receptor gene bodies. Each row is a genomic region, each column is a
sample.

```{r}
h3k9me3 <- read_tsv(
  file.path(chipseq_dir, "h3k9me3_signal", "all_H3K9me3_signal_mm10.tab"),
  col_names = TRUE
)

# Clean column names — remove quotes and file extensions
colnames(h3k9me3) <- colnames(h3k9me3) |>
  str_remove_all("'") |>
  str_remove("\\.mLb\\.clN\\.bigWig")

# Rename for clarity
h3k9me3 <- h3k9me3 |>
  rename(
    chr   = `#chr`,
    start = start,
    end   = end,
    Bulk  = ChIP_Bulk_mOSN_REP1,
    Z1_R1 = ChIP_Zone1_mOSN_REP1,
    Z1_R2 = ChIP_Zone1_mOSN_REP2,
    Z2_R1 = ChIP_Zone2_mOSN_REP1,
    Z2_R2 = ChIP_Zone2_mOSN_REP2,
    Z5_R1 = ChIP_Zone5_mOSN_REP1,
    Z5_R2 = ChIP_Zone5_mOSN_REP2
  ) |>
  mutate(
    Zone1_signal = (Z1_R1 + Z1_R2) / 2,
    Zone2_signal = (Z2_R1 + Z2_R2) / 2,
    Zone5_signal = (Z5_R1 + Z5_R2) / 2
  )

dim(h3k9me3)
head(h3k9me3)
```

## 2.2 Load olfactory receptor BED files

```{r}
# BED file columns: chr, start, end, gene_name, score, strand
bed_cols <- c("chr", "start", "end", "gene_name", "score", "strand")

or_zone1 <- read_tsv(
  file.path(chipseq_dir, "or_bed", "OR_genes_zone1_mm10.bed"),
  col_names = bed_cols
)

or_zone2 <- read_tsv(
  file.path(chipseq_dir, "or_bed", "OR_genes_zone2_mm10.bed"),
  col_names = bed_cols
)

or_zone5 <- read_tsv(
  file.path(chipseq_dir, "or_bed", "OR_genes_zone5_mm10.bed"),
  col_names = bed_cols
)

or_all <- read_tsv(
  file.path(chipseq_dir, "or_bed", "OR_genes_zone_all_mm10.bed"),
  col_names = bed_cols
)

# Add zone label
or_zone1 <- or_zone1 |> mutate(expressed_zone = "Zone 1")
or_zone2 <- or_zone2 |> mutate(expressed_zone = "Zone 2")
or_zone5 <- or_zone5 |> mutate(expressed_zone = "Zone 5")

# How many OR genes per zone?
cat("Zone 1:", nrow(or_zone1), "OR genes\n")
cat("Zone 2:", nrow(or_zone2), "OR genes\n")
cat("Zone 5:", nrow(or_zone5), "OR genes\n")
```

## 2.3 Load RNA-seq DESeq2 results

```{r}
# Load DESeq2 object (all 8 samples, all three zones)
dds_all <- readRDS(here("analysis", "solutions", "dds_all.rds"))

# Load gene annotation
gene_info <- readRDS(here("analysis", "solutions", "gene_info.rds"))

# Load pairwise DESeq2 results tables
res_z1_z2 <- read_csv(here("analysis", "solutions", "results_zone1_vs_zone2.csv"))
res_z1_z5 <- read_csv(here("analysis", "solutions", "results_zone1_vs_zone5.csv"))
res_z2_z5 <- read_csv(here("analysis", "solutions", "results_zone2_vs_zone5.csv"))

# Quick check
dds_all
head(res_z1_z2)
```

---

# 3. H3K9me3 Signal at Olfactory Receptor Loci

## 3.1 Join H3K9me3 signal with OR gene coordinates

The H3K9me3 signal table contains coordinates — we join it with the BED files
by genomic position to assign signal values to specific OR genes.

```{r}
# Join H3K9me3 signal with OR gene coordinates by genomic position
join_signal <- function(h3k9me3, or_bed) {
  or_bed |>
    left_join(h3k9me3, by = c("chr", "start", "end")) |>
    filter(!is.na(Zone1_signal))
}

or_z1_signal <- join_signal(h3k9me3, or_zone1)
or_z2_signal <- join_signal(h3k9me3, or_zone2)
or_z5_signal <- join_signal(h3k9me3, or_zone5)

# Combine all
or_all_signal <- bind_rows(or_z1_signal, or_z2_signal, or_z5_signal)

head(or_all_signal)
```

## 3.2 H3K9me3 signal by expressed zone

Compare H3K9me3 levels in each zone for OR genes expressed in Zone 1, Zone 2, or Zone 5.

```{r}
# Reshape to long format for plotting
or_signal_long <- or_all_signal |>
  dplyr::select(gene_name, expressed_zone, Zone1_signal, Zone2_signal, Zone5_signal) |>
  pivot_longer(
    cols      = c(Zone1_signal, Zone2_signal, Zone5_signal),
    names_to  = "measured_zone",
    values_to = "h3k9me3_signal"
  ) |>
  mutate(
    measured_zone = str_remove(measured_zone, "_signal") |>
      str_replace("Zone", "Zone ")
  )

# Boxplot: H3K9me3 signal in each zone for different OR gene groups
ggplot(or_signal_long,
       aes(x     = measured_zone,
           y     = h3k9me3_signal,
           fill  = measured_zone)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  geom_jitter(width = 0.1, size = 0.3, alpha = 0.3) +
  facet_wrap(~ expressed_zone,
             labeller = labeller(expressed_zone = function(x) paste("OR genes expressed in", x))) +
  scale_fill_manual(values = c(
    "Zone 1" = "#E41A1C",
    "Zone 2" = "#377EB8",
    "Zone 5" = "#4DAF4A"
  )) +
  theme_bw() +
  labs(
    title = "H3K9me3 signal at olfactory receptor loci",
    x     = "Zone measured",
    y     = "H3K9me3 signal",
    fill  = "Zone"
  ) +
  theme(
    plot.title      = element_text(hjust = 0.5),
    legend.position = "none",
    strip.text      = element_text(size = 9)
  )
```

> **Question:** For OR genes expressed in Zone 1, what do you expect to see in Zone 2
> and Zone 5? Does the plot support the hypothesis that H3K9me3 marks silent OR genes?

---

# 4. Correlating H3K9me3 with Gene Expression

## 4.1 Prepare the joined dataset

To correlate ChIP-seq signal with RNA-seq fold changes, we need to link genes
between the two datasets. The OR BED files contain gene names (Olfr symbols),
and the RNA-seq results contain Ensembl IDs — we use gene names to join them.

```{r}
# Join RNA-seq results with H3K9me3 signal for Zone 1 vs Zone 2
# We look for OR genes that are DE between zones

# Get OR gene names from all zones
all_or_genes <- or_all |> pull(gene_name) |> unique()

# Filter RNA-seq results to OR genes only
res_or_z1_z2 <- res_z1_z2 |>
  filter(external_gene_name %in% all_or_genes)

res_or_z1_z5 <- res_z1_z5 |>
  filter(external_gene_name %in% all_or_genes)

res_or_z2_z5 <- res_z2_z5 |>
  filter(external_gene_name %in% all_or_genes)


cat("OR genes in zone1_vs_zone2 results:", nrow(res_or_z1_z2), "\n")
cat("OR genes in zone1_vs_zone5 results:", nrow(res_or_z1_z5), "\n")
cat("OR genes in zone2_vs_zone5 results:", nrow(res_or_z2_z5), "\n")
```

## 4.2 Scatter plot: log2FC vs H3K9me3 signal difference

For Zone 1 vs Zone 2: genes upregulated in Zone 1 (positive log2FC) should show
**lower** H3K9me3 in Zone 1 and **higher** H3K9me3 in Zone 2.

```{r}
# Compute H3K9me3 signal difference (Zone1 - Zone2) for OR genes
h3k9me3_diff_z1z2 <- or_all_signal |>
  dplyr::select(gene_name, Zone1_signal, Zone2_signal, Zone5_signal) |>
  mutate(diff_z1_z2 = Zone1_signal - Zone2_signal,
         diff_z1_z5 = Zone1_signal - Zone5_signal,
         diff_z2_z5 = Zone2_signal - Zone5_signal)

# Join with RNA-seq results
cor_z1_z2 <- res_or_z1_z2 |>
  left_join(h3k9me3_diff_z1z2, by = c("external_gene_name" = "gene_name")) |>
  filter(!is.na(log2FoldChange), !is.na(diff_z1_z2))

# Label top DE OR genes
top_or <- cor_z1_z2 |>
  filter(!is.na(padj)) |>
  arrange(padj) |>
  head(15)

ggplot(cor_z1_z2,
       aes(x = log2FoldChange,
           y = diff_z1_z2)) +
  geom_point(aes(colour = case_when(
    !is.na(padj) & padj < 0.05 & log2FoldChange >  1 ~ "Up in Zone 1",
    !is.na(padj) & padj < 0.05 & log2FoldChange < -1 ~ "Up in Zone 2",
    TRUE ~ "Not significant"
  )), alpha = 0.6, size = 1.5) +
  geom_text_repel(
    data         = top_or,
    aes(label    = external_gene_name),
    size         = 3,
    max.overlaps = 15,
    colour       = "black"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c(
    "Up in Zone 1"    = "#E41A1C",
    "Up in Zone 2"    = "#377EB8",
    "Not significant" = "grey70"
  )) +
  theme_bw() +
  labs(
    title   = "Zone 1 vs Zone 2: RNA-seq log2FC vs H3K9me3 signal difference",
    x       = "Log2 fold change (RNA-seq)\nPositive = higher in Zone 1",
    y       = "H3K9me3 signal difference (Zone1 - Zone2)\nPositive = more H3K9me3 in Zone 1",
    colour  = NULL
  ) +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 11),
    legend.position = "top"
  )
```

> **Question:** What quadrant would you expect most Zone 1-expressed OR genes to
> fall in? (Hint: if a gene is expressed in Zone 1 and silenced in Zone 2, what
> happens to its RNA-seq fold change AND its H3K9me3 signal?)

> **Exercise:** Repeat the scatter plot and correlation for Zone 1 vs Zone 5 and
> Zone 2 vs Zone 5.

---

# 6. Zone-specific expression vs silencing

## 6.1 Compare H3K9me3 in "own zone" vs "other zones"

For each OR gene, compare H3K9me3 signal in its expressed zone vs the other zones.

```{r}
# For each OR gene, extract signal in "own zone" and "other zones"
own_vs_other <- or_all_signal |>
  mutate(
    own_zone_signal = case_when(
      expressed_zone == "Zone 1" ~ Zone1_signal,
      expressed_zone == "Zone 2" ~ Zone2_signal,
      expressed_zone == "Zone 5" ~ Zone5_signal
    ),
    other_zone_signal = case_when(
      expressed_zone == "Zone 1" ~ (Zone2_signal + Zone5_signal) / 2,
      expressed_zone == "Zone 2" ~ (Zone1_signal + Zone5_signal) / 2,
      expressed_zone == "Zone 5" ~ (Zone1_signal + Zone2_signal) / 2
    ),
    signal_difference = other_zone_signal - own_zone_signal
  )

# Boxplot: own zone vs other zones
own_vs_other |>
  dplyr::select(gene_name, expressed_zone, own_zone_signal, other_zone_signal) |>
  pivot_longer(
    cols      = c(own_zone_signal, other_zone_signal),
    names_to  = "signal_type",
    values_to = "signal"
  ) |>
  mutate(signal_type = recode(signal_type,
    "own_zone_signal"   = "Expressed zone",
    "other_zone_signal" = "Silent zones (mean)"
  )) |>
  ggplot(aes(x = signal_type, y = signal, fill = signal_type)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  geom_jitter(width = 0.1, size = 0.3, alpha = 0.3) +
  facet_wrap(~ expressed_zone) +
  scale_fill_manual(values = c(
    "Expressed zone"       = "#66C2A5",
    "Silent zones (mean)"  = "#FC8D62"
  )) +
  theme_bw() +
  labs(
    title = "H3K9me3 signal: expressed zone vs silent zones",
    x     = NULL,
    y     = "H3K9me3 signal",
    fill  = NULL
  ) +
  theme(
    plot.title      = element_text(hjust = 0.5),
    legend.position = "top"
  )
```

---

# Biological Conclusions


1. **Does H3K9me3 mark silent olfactory receptors?** Is the signal consistently
   lower at OR genes in their expressed zone compared to silent zones?

2. **Does the pattern hold for all zones equally?** Zone 5 has only 21 expressed
   OR genes — does the smaller sample size affect your confidence in the results?

3. **Limitations of this analysis:** What assumptions are we making? What
   additional data would strengthen these conclusions?

---

*By Lisa Korneeva with some help of Claude Sonnet 4.6*
